# A Conflict Inflation Model

## Overview

This model is a version of the conflicting-claims approach to inflation
that has been developed in the Marxian and post-Keynesian tradition
(@Rowthorn1977).[^a_conflict_inflation_model-1] The conflict inflation
approach argues that inflation is driven by distributional conflict
between price-setting firms and wage-setting workers. Each social class
aspires a share of real income that is inconsistent with the claims of
the other class. Workers gradually adjust nominal wages to reach a real
wage target. Firms, by contrast, gradually adjust prices to achieve an
implicit real wage target. The (implicit) real wage target of firms may
stem from an (explicit) profitability target. Profitability, in turn,
depends on the unit cost of firms. With nominal wages being a key
component of those unit cost, any attempt by workers to raise real wages
through higher nominal wages will eat into firms' profitability, which
will respond by raising prices. Equally, any attempt by firms to raise
profitability by setting higher prices will undermine real wages and
evoke a response by workers. This interplay between mutually
inconsistent claims drives the inflationary process and the equilibrium
real wage. Distributional outcomes in this model depend on the
distributional targets of workers and firms as well as their ability to
adjust wages and prices, which in turn may be affected by their relative
bargaining power and labour market institutions.

[^a_conflict_inflation_model-1]: With textbook treatments in
    @Blecker2019, chap. 5, @Hein2023, chap. 5, and @Lavoie2022, chap. 8.

We present a simplified version of the [three-sector conflict inflation
model](https://blogs.lse.ac.uk/businessreview/2023/11/29/how-to-reduce-inflation-without-increasing-unemployment-and-income-inequality/) in @Wildauer2023 that examines the inflationary and distributional
effects of energy price shocks.[^a_conflict_inflation_model-2] A final
output sector uses energy as an intermediate input. Exogenous shocks to
the prices of energy inputs can trigger temporary 'price-wage spirals',
where firms raise prices to defend their profit margins, which in turn
induces workers to raise nominal wages, which prolongs the inflationary
shocks. While real wages fall, the economy-wide profit share rises.

[^a_conflict_inflation_model-2]: @Wildauer2023 use this model to study
    the 2021-23 inflationary episode in the US. Besides an energy sector
    providing intermediate inputs, they consider a final goods and a
    final services sector. In addition to the energy price shock, there
    is a positive demand shift towards goods and a negative supply shock
    due to lockdowns, creating a rise in the rate of capacity
    utilisation of the final goods sector. These additional shocks allow
    the final goods sector to raise its markup. The present simplified
    version of the model only considers the energy sector and a final
    output sector, and treats the markup of the final output sector as
    constant.

## The Model

The model is given by the following equations:

$$
\Omega_t = \frac{w_t}{p_t} 
$$ {#eq-conflict_omega} $$
w_{t}  = w_{t-1}\left[ 1 + \phi(\Omega^W - \Omega_{t-1}) \right]
$$ {#eq-conflict_w}

$$
p_t= p_{t-1} \left[ 1 + \psi (r^T - r_{t-1}) \right] 
$$ {#eq-conflict_p}

$$
p^T_t=(1+\theta^T)\left(w_t + \delta p^E\right)  
$$ {#eq-conflict_pf}

$$
r^T=\frac{\theta^T}{1+\theta^T}
$$ {#eq-conflict_rf}

$$
r_T=1-\Omega_t - \delta \epsilon
$$ {#eq-conflict_r} $$
p^E_t = \epsilon p_t 
$$ {#eq-conflict_p_E}

where $\Omega$, $w$, $p$, $\Omega^W$, $r^T$, $r$, $p^T$, and $p^E$
are the real wage, the nominal wage, the price level, workers' desired
real wage, firms' desired profit margin, the realised profit margin,
firms' desired price level, and the nominal energy price, respectively.

@eq-conflict_omega defines the real wage $\Omega$. @eq-conflict_w
specifies the dynamics of wage setting. Workers set the growth rate of
nominal wages $w$ proportionally to the gap between their target real
wage $\Omega^W$ and the actual real wage $\Omega$. The target real wage
of workers is taken to be exogenous in this version of the
model.[^a_conflict_inflation_model-3] By @eq-conflict_p, firms set the
price inflation rate in a symmetric manner, responding to the deviation
of the desired profit margin $r^T$ from the actual profit margin $r$.
@eq-conflict_pf specifies the desired price level of firms, which charge
a desired markup $\theta^T$ on their unit variable cost. Unit costs in this
model consist of two components: (i) unit labour cost given by $w_t$
(where we normalise labour productivity to unity for simplicity such
that $\frac{Y}{L}=1$), and (ii) unit energy cost $\delta p^E$, where
$\delta =\frac{E}{Y}$ is the (exogenous) energy intensity and $p^E$
is the nominal energy price. The profit margin is defined as profits
over gross output:
$r=\frac{\Pi}{pY}=\frac{pY-wL-p^E E}{pY}$.
The *target* profit margin as implied by the desired price level is then
$r^T = \frac{p^T - (w_t + \delta p^E) }{p^T}$, yielding
@eq-conflict_rf. The desired markup $\theta^T$ is considered exogenous in
this version of the model.[^a_conflict_inflation_model-4] The actual or
realised profit margin is given by @eq-conflict_r, where
$\epsilon=\frac{p^E}{p}$ is the real energy price. Finally, by
@eq-conflict_p_E, the real energy price $\epsilon$ is taken to be
exogenous (and subject to shocks), implying that the nominal energy
price must grow at the same rate as the price level.

[^a_conflict_inflation_model-3]: @Blecker2019, chap. 5 and @Lavoie2022,
    chap. 8 analyse extensions which endogenise the real wage target of
    workers.

[^a_conflict_inflation_model-4]: In @Wildauer2023, the target markup of
    the final goods sector is endogenous with respect to the rate of
    capacity utilisation.


## Simulation

### Parameterisation

Table 1 reports the parameterisation used in the simulation. Besides a
baseline (labelled as scenario 1), three further scenarios will be
considered. In scenario 2, the real energy price $\epsilon$ increases.
In scenario 3, the target markup $\theta^T$ rises. In scenario 4, the target
real wage $\Omega^W$ rises.

**Table 1: Parameterisation**

| Scenario                                  | $\phi$ | $\psi$ | $\delta$ | $\epsilon$ | $\theta^T$ | $\Omega^W$ |
|--------------------|---------|---------|---------|---------|---------|---------|
| 1: baseline                               | 1      | 1        | 1        | 0.55       | 0.4   | 0.2        |
| 2: rise in real energy price $(\epsilon)$ | 1      | 1        | 1        | 0.8        | 0.4   | 0.2        |
| 3: rise in target markup $(\theta^T)$          | 1      | 1        | 1        | 0.55       | 0.6   | 0.2        |
| 4: rise in target real wage $(\Omega^W)$  | 1      | 1        | 1        | 0.55       | 0.4   | 0.3        |

### Simulation code

```{r conflict1}

#Clear the environment
rm(list=ls(all=TRUE))

#Set number of periods
Q = 300

# Set number of scenarios (including baselines)
S=4

# Set period in which exogenous shift will occur
q=50

#Create (S x Q) matrices in which equilibrium solutions from different parameterisations will be stored
omega=matrix(data=0.5, nrow=S, ncol=Q)   # real wage
w=matrix(data=0.1, nrow=S, ncol=Q)       # nominal wage
p=matrix(data=0.5, nrow=S, ncol=Q)       # price level
w_hat=matrix(data=0.1, nrow=S, ncol=Q)   # growth rate of nominal wages
p_hat=matrix(data=0.1, nrow=S, ncol=Q)   # growth rate of prices = inflation rate
p_T=matrix(data=0.5, nrow=S, ncol=Q)     # firms' desired price level
omega_f=matrix(data=0.5, nrow=S, ncol=Q) # firms' desired real wage 
p_E=matrix(data=0.5, nrow=S, ncol=Q)   # nominal energy price
r_T=matrix(data=0.5, nrow=S, ncol=Q)     # target profit margin
r=matrix(data=0.5, nrow=S, ncol=Q)       # realised profit margin

# Set constant parameter values
phi=1     # adjustment speed of nominal wages
psi=1   # adjustment speed of prices
delta=1   # energy intensity

# Set and initialise exogenous variables/parameters that will be shifted
epsilon=matrix(data=0.55, nrow=S, ncol=Q)  # real energy price
theta_T=matrix(data=0.4, nrow=S, ncol=Q)       # target markup
omega_w=matrix(data=0.2, nrow=S, ncol=Q)   # target real wage of workers

# Set parameter values for different scenarios
epsilon[2,q:Q]=0.8 # scenario 2: rise in real energy price
theta_T[3,q:Q]=0.6     # scenario 3: rise in target markup
omega_w[4,q:Q]=0.3 # scenario 4: rise in target real wage of workers

# Simulate the model by looping over Q time periods for S different scenarios
for (i in 1:S){
    
  for (t in 2:Q){
    
    for (iterations in 1:1000){ # iterate the model 1000-times in each period
    
    #(1) Real wage
    omega[i,t] =w[i,t]/p[i,t]
      
    #(2) Nominal wage
    w[i,t] = w[i,t-1]*(1 + phi*(omega_w[i,t-1] - omega[i,t-1]))
    
    #(3) Price level
    p[i,t] = p[i, t-1]*(1 + psi*(r_T[i, t-1] - r[i, t-1]))
    
    #(4) Firms' desired real wage
    omega_f[i,t] = w[i,t]/p_T[i,t]
    
    #(5) Firms' desired price level
    p_T[i,t] = (1 + theta_T[i,t])*(w[i,t] + delta * p_E[i,t])
    
    #(6) Nominal energy price
    p_E[i,t]=epsilon[i,t]*p[i,t]
    
    # Target profit margin
    r_T[i, t-1] = theta_T[i, t]/(1 + theta_T[i, t])
    
    # Realised profit margin
    r[i, t] = 1 - omega[i, t] - epsilon[i,t]*delta
    
    # Growth rate of nominal wages
    w_hat[i,t]= (w[i,t]- w[i,t-1])/w[i,t-1]
    
    # Growth rate of prices = inflation rate
    p_hat[i,t]= (p[i,t]- p[i,t-1])/p[i,t-1]

  } # close iterations loop
 }  # close time loop
}   # close scenarios loop

```

::: {.callout-note collapse="true" title="Python code"}
```{python}
#| echo: true
#| output: false
#| collapse: true

import numpy as np

# Set number of periods
Q = 300

# Set number of scenarios (including baseline)
S = 4

# Set period in which exogenous shift will occur
q = 50

# Create (S x Q) matrices in which equilibrium solutions from different parameterizations will be stored
omega = np.full((S, Q), 0.5)  # real wage
w = np.full((S, Q), 0.1)      # nominal wage
p = np.full((S, Q), 0.5)      # price level
w_hat = np.full((S, Q), 0.1)  # growth rate of nominal wages
p_hat = np.full((S, Q), 0.1)  # growth rate of prices = inflation rate
p_T = np.full((S, Q), 0.5)    # firms' desired price level
omega_f = np.full((S, Q), 0.5) # firms' desired real wage 
p_E = np.full((S, Q), 0.5)  # nominal energy price
r_T = np.full((S, Q), 0.5)    # target profit margin
r = np.full((S, Q), 0.5)      # realised profit margin

# Set constant parameter values
phi = 1     # adjustment speed of nominal wages
psi = 1   # adjustment speed of prices
delta = 1   # energy intensity

# Set and initialize exogenous variables/parameters that will be shifted
epsilon = np.full((S, Q), 0.55)  # real energy price
theta_T = np.full((S, Q), 0.4)       # target markup
omega_w = np.full((S, Q), 0.2)   # target real wage of workers

# Set parameter values for different scenarios
epsilon[1, q:Q] = 0.8  # scenario 2: rise in real energy price
theta_T[2, q:Q] = 0.6      # scenario 3: rise in target markup
omega_w[3, q:Q] = 0.3  # scenario 4: rise in target real wage of workers

# Simulate the model by looping over Q time periods for S different scenarios
for i in range(S):
    for t in range(1, Q):
        for _ in range(1000):  # iterate the model 1000-times in each period
            
            # (1) Real wage
            omega[i, t] = w[i, t] / p[i, t]
              
            # (2) Nominal wage
            w[i, t] = w[i, t-1] * (1 + phi * (omega_w[i, t-1] - omega[i, t-1]))
            
            # (3) Price level
            p[i, t] = p[i, t-1] * (1 + psi * (r_T[i, t-1] - r[i, t-1]))
            
            # (4) Firms' desired real wage
            omega_f[i, t] = w[i, t] / p_T[i, t]
            
            # (5) Firms' desired price level
            p_T[i, t] = (1 + theta_T[i, t]) * (w[i, t] + delta * p_E[i, t])
            
            # (6) Nominal energy price
            p_E[i, t] = epsilon[i, t] * p[i, t]
            
            # Target profit margin
            r_T[i, t-1] = theta_T[i, t] / (1 + theta_T[i, t])
            
            # Realised profit margin
            r[i, t] = 1 - omega[i, t] - epsilon[i, t] * delta
            
            # Growth rate of nominal wages
            w_hat[i, t] = (w[i, t] - w[i, t-1]) / w[i, t-1]
            
            # Growth rate of prices = inflation rate
            p_hat[i, t] = (p[i, t] - p[i, t-1]) / p[i, t-1]
            
```
:::

### Plots

@fig-baseline-prices displays the dynamics of nominal wages and prices
(in levels), both of which exhibit exponential growth.
@fig-baseline-inflation plots the corresponding growth rates: the price
and wage inflation rates. It can be seen that after an initial
adjustment period, both rates converge on the same value, implying a
constant real wage (and a constant profit margin).

```{r conflict_2}
#| label: fig-baseline-prices
#| fig-cap: Prices and wages (baseline)

# Set start and end periods for plots
Tmin =3
Tmax=300

# Prices and wages (baseline)
plot(p[1, Tmin:Tmax],type="l", col=1, lwd=2, lty=1, xlab="Time", ylab="p, w") 
title(main="Prices and wages (baseline)", cex=0.8)
lines(w[1, Tmin:Tmax],lty=2)
legend("topleft", legend=c("p", "w"),
       lty=1:2, cex=0.8, bty = "n", y.intersp=0.8)

```

```{r conflict_3}
#| label: fig-baseline-inflation
#| fig-cap: Price and wage inflation (baseline)

# Set start and end periods for plots
Tmin =3
Tmax=60

# Price and wage inflation (in %) (baseline)
plot(100*p_hat[1, Tmin:Tmax],type="l", col=1, lwd=2, lty=1, xlab="Time", ylab="%", ylim=range(100*p_hat[1, Tmin:Tmax], 100*w_hat[1, Tmin:Tmax])) 
title(main="Price and wage inflation (baseline)", cex=0.8)
lines(100*w_hat[1, Tmin:Tmax],lty=2)
legend("topright", legend=c(expression(hat(p)), expression(hat(w))),
       lty=1:2, cex=0.8, bty = "n", y.intersp=0.8)
```

@fig-energy displays the energy price shock (scenario 2), where the real
energy price rises permanently.[^a_conflict_inflation_model-5] As a
result, there is a bout in the inflation rate, followed by an increase
in the wage inflation rate. This reaction may be characterised as a
(temporary) 'price-wage spiral' (@Wildauer2023), where firms react to
the cost shock by raising prices to protect their profit margins, which
trigger a response by workers to defend their real wages, which prolongs
the inflationary shock. However, the inflation rate then embarks on a
gradual downward convergence towards a new equilibrium that is higher
than the one before the shock. Since the price inflation rate exceeds
the wage inflation rate during the adjustment process, the real wage
falls in response to the increase in the real energy price.

[^a_conflict_inflation_model-5]: More accurately this is an energy price
    'shift'.

```{r conflict_4}
#| label: fig-energy
#| fig-cap: Price and wage inflation and real wage (energy price shock)

# Set start and end periods for plots
Tmin =10
Tmax=100

# Inflation and real wage in energy price shock scenario
plot(100*p_hat[2, Tmin:Tmax], type="l", col=1, lwd=2, lty=1, xlab="Time", ylab=expression(hat(p))) 
title(main="Wage and price inflation and real wage, energy price shock (scen. 2)", cex=0.8)
lines(100*w_hat[2, Tmin:Tmax],lty=2, lwd=2)
mtext(expression(hat(w)), side = 2, line = 3, at=12)
par(mar = c(5, 4, 4, 4) + 0.3)
par(new = TRUE)
plot(omega[2, Tmin:Tmax], type="l", col=1, lwd=2, lty=3, font.main=1, cex.main=1,ylab = '', axes=FALSE,
     xlab = '', ylim = range(omega[2, Tmin:Tmax]), cex=0.8)
axis(side = 4, at=pretty(omega[2, Tmin:Tmax]))  
mtext(expression(Omega), side = 4, line = 3)
legend("topright", legend=c(expression(hat(p)),expression(hat(w)), expression(Omega)),
       lty=1:3, cex=0.8, bty = "n", y.intersp=0.8)
```


@fig-markup considers the scenario where the target markup of firms
rises. The resulting dynamics are very similar to the energy price
shock: the inflation rate spikes, triggering a response in nominal wages
that is weaker than the increase in inflation. As a result, the real
wage falls until a new equilibrium with a higher inflation rate has been
reached. In contrast to the energy price shock, both the profit margin
and the profit share must increase in this scenario.

```{r conflict_6}
#| label: fig-markup
#| fig-cap: Increase in target markup

# Set start and end periods for plots
Tmin =30
Tmax=80

# Inflation and real wage in scenario 3
plot(100*p_hat[3, Tmin:Tmax], type="l", col=1, lwd=2, lty=1, xlab="Time", ylab=expression(hat(p))) 
title(main="Price and wage inflation and real wage, rise in target markup (scen. 3)", cex=0.8)
lines(100*w_hat[3, Tmin:Tmax],lty=2, lwd=2)
mtext(expression(hat(w)), side = 2, line = 3, at=7)
par(mar = c(5, 4, 4, 4) + 0.3)
par(new = TRUE)
plot(omega[3, Tmin:Tmax], type="l", col=1, lwd=2, lty=3, font.main=1, cex.main=1,ylab = '', axes=FALSE,
     xlab = '', ylim = range(omega[3, Tmin:Tmax]), cex=0.8)
axis(side = 4, at=pretty(omega[3, Tmin:Tmax]))  
mtext(expression(Omega), side = 4, line = 3)
legend("topright", legend=c(expression(hat(p)),expression(hat(w)), expression(Omega)),
       lty=1:3, cex=0.8, bty = "n", y.intersp=0.8)
```

@fig-targetrealwage displays scenario 4, where the target real wage of
workers rises. The wage inflation rate shoots up, which is followed by
an increase in price inflation. As nominal wages temporarily grow faster
than prices, the real wage increases. By implication, both profit share
and margin will fall.

Taken together, scenarios 3 and 4 illustrate that besides cost shocks,
exogeneous changes in the price- and wage-setting behaviour of firm and
workers, respectively, will change the equilibrium inflation rate as
well as income distribution. Such changes could stem from increases in
market power (scenario 3) or fiercer industrial action by workers
(scenario 4). An increase of the claim of either party on total real
output will result in elevated social conflict, reflected in an increase
in the inflation rate. The resulting distributional effect will depend
on which party initiates the conflict.

```{r conflict_7}
#| label: fig-targetrealwage
#| fig-cap: Increase in target real wage of workers

# Inflation and real wage in scenario 4
plot(100*p_hat[4, Tmin:Tmax], type="l", col=1, lwd=2, lty=1, xlab="Time", ylab=expression(hat(p)), ylim=range(100*p_hat[4, Tmin:Tmax], 100*w_hat[4, Tmin:Tmax])) 
title(main="Price and wage inflation real wage, rise in target real wage (scen. 4)", cex=0.8)
lines(100*w_hat[4, Tmin:Tmax],lty=2, lwd=2)
mtext(expression(hat(w)), side = 2, line = 3, at=7.5)
par(mar = c(5, 4, 4, 4) + 0.3)
par(new = TRUE)
plot(omega[4, Tmin:Tmax], type="l", col=1, lwd=2, lty=3, font.main=1, cex.main=1,ylab = '', axes=FALSE,
     xlab = '', ylim = range(omega[4, Tmin:Tmax]), cex=0.8)
axis(side = 4, at=pretty(omega[4, Tmin:Tmax]))  
mtext(expression(Omega), side = 4, line = 3)
legend("topright", legend=c(expression(hat(p)),expression(hat(w)), expression(Omega)),
       lty=1:3, cex=0.8, bty = "n", y.intersp=0.8)
```

::: {.callout-note collapse="true" title="Python code"}
```{python}
#| echo: true
#| output: false
#| collapse: true

## Plot (here only inflation in scenario 2)            
import matplotlib.pyplot as plt

# Set start and end periods for plots
Tmin = 9   # Zero-based index (equivalent to Tmin=10 in R)
Tmax = 100

# Set up the main plot
fig, ax1 = plt.subplots()

# Plot inflation rates (price and wage growth rates) on the primary y-axis
ax1.plot(100 * p_hat[1, Tmin:Tmax], label=r'$\hat{p}$', color="black", linewidth=2, linestyle="-")  # price inflation
ax1.plot(100 * w_hat[1, Tmin:Tmax], label=r'$\hat{w}$', color="black", linewidth=2, linestyle="--")  # wage inflation
ax1.set_xlabel("Time")
ax1.set_ylabel(r'$\hat{p}$, $\hat{w}$')
ax1.set_ylim(0, 30)  # Adjust this limit as needed based on your data

# Add secondary y-axis for real wage
ax2 = ax1.twinx()
ax2.plot(omega[1, Tmin:Tmax], label=r'$\Omega$', color="black", linewidth=2, linestyle=":")  # real wage
ax2.set_ylabel(r'$\Omega$')
ax2.set_ylim(0, 0.2)  # Set range based on omega values

# Titles and legends
plt.title("Wage and price inflation and real wage, energy price shock (scenario 2)", fontsize=10)
ax1.legend(loc="upper left", frameon=False, fontsize=8)
ax2.legend(loc="upper right", frameon=False, fontsize=8)

plt.show()
            
```
:::

## A closer look at the effect of energy price shocks on profits

The model can be used to study the effects of an energy price shock on
different measures of profitability and income distribution. Table 2 provides an overview of
these measures.

**Table 2: Measures of profits**

|                                               | Definition                                 | Model equation                                                                                | Effect of energy price shock^[We derive the exact effects in the analytical discussion below.] $(\uparrow \epsilon)$ |
|---------------|---------------|----------------|---------------|
| Profit margin = profit share of energy sector | $$ r^E = \pi^E = \frac{\Pi^E}{p^E E} $$ | $$ r^E = \frac{p^E E - w^E L^E}{p^E E} = 1 - \frac{\Omega^E}{\epsilon} $$               | Positive                                                   |
| Profit margin of final output sector          | $$ r = \frac{\Pi}{pY} $$                   | $$ r = \frac{pY - wL - p^E E}{pY} = 1 - \Omega - \delta \epsilon $$                        | Negative                                                   |
| Profit share of final output sector           | $$ \pi = \frac{\Pi}{pVA} $$                | $$ \pi = \frac{pY - wL - p^E E}{pY - p^E E} = 1 - \frac{\Omega}{1 - p^E \epsilon} $$ | Ambiguous                                                  |
| Total profit share                            | $$ \pi^{Total} = \frac{\Pi + \Pi^E}{pY} $$ | $$ \pi^{Total} = \frac{pY - w^E L^E - wL}{pY} = 1 - \delta \Omega^E - \Omega $$               | Positive                                                   |

**Note:** $\Pi$ represents nominal profits, $E$ is the output of
the energy sector, which is an input to the final output
$Y$. The real wage of the energy sector $\Omega^E = w^E/p$ is
exogenous. Labour productivity in both sectors is normalised to unity:
$L^E / E = L / Y = 1$. $VA$ is value added.

The profit *margin* $r$ is defined as the share of
profits in *gross* output. By contrast, the profit *share* $\pi$ is
defined as the share of profits in *value added*, i.e. output net of
intermediate inputs. For the energy sector, which in this model does not use any intermediate inputs, the profit margin and profit share coincide. With the real wage in the energy sector assumed exogenous, a real energy price increase will unambiguously raise the profits of the energy sector.
For the final output sector that uses energy as an input, an energy price shock will lower the profit margin due to the increase in energy cost. However, the effect on the final output sector's profit share are ambiguous and will depend on the relative ability of workers and firms to defend their income targets in response to the shock. Finally, the total profit share incorporates the profits of the energy input sector and the final output sector (this assumes that the energy sector is domestic). An increase in the real energy price has an unambiguously positive effect on the economy-wide profit share. 

In @fig-energy-prof, it can be seen how the energy price shock affects
the profit margin and profit share of the final output sector in the simulations. While the profit margin *falls* due
to the increase in energy cost, the profit share *rises*. This is because for the chosen parameterisation, we have $\psi r^T-\phi\Omega^W>0$, for which $\frac{\partial \pi}{\partial \epsilon} > 0$.^[See analytical discussion below.] Thus, if the price-setting
power of firms and their target profit margins $\psi r^T$ are sufficiently large  relative to wage-setting power and  real wage target of workers $\phi\Omega^W$, profit shares of domestic non-energy sectors may rise in response to an energy cost shock. 

```{r conflict_5}
#| label: fig-energy-prof
#| fig-cap: Profit margin and profit share (energy price shock)

# Calculate final output profit share
pi=(p-w-p_E*delta)/(p-p_E*delta)

# Check if psi*r_T > phi*omega_w, implying that the effect of an energy price increase on the profit share of the final output sector is positive
print(psi*r_T[1,1] - phi*omega_w[1,1] >0)

# Plot profit margin and profit share in energy price shock scenario
plot(r[2, Tmin:Tmax], type="l", col=1, lwd=2, lty=1, xlab="Time", ylab="r,", ylim=range(r[2, Tmin:Tmax], pi[2, Tmin:Tmax])) 
title(main="Profit margin vs profit share, energy price shock (scenario 2)", cex=0.8)
lines(pi[2, Tmin:Tmax],lty=2, lwd=2)
mtext(expression(pi), side = 2, line = 3, at=0.4)
legend("right", legend=c("r", expression(pi)),
       lty=1:2, cex=0.8, bty = "n", y.intersp=0.8)
```

## Directed graph

Another perspective on the model's properties is provided by its
directed graph. A directed graph consists of a set of nodes that
represent the variables of the model. Nodes are connected by directed
edges. An edge directed from a node $x_1$ to node $x_2$ indicates a
causal impact of $x_1$ on $x_2$.

```{r conflict_8, message=FALSE}
## Create directed graph
# Construct auxiliary Jacobian matrix for 9 variables: 
  # endogenous: (1) Omega, (2) w, (3) p, (4) r^T, (5) r, (6) p_E
  # exogenous: (7) epsilon, (8) \theta^T, (9) Omega^W

              #O w p rF r g e mF OW
M_mat=matrix(c(0,1,1, 0,0,0,0, 0, 0,  # Omega
               1,0,0, 0,0,0,0, 0, 1,  # w
               0,0,0, 1,1,0,0, 0, 0,  # p
               0,0,0, 0,0,0,0, 1, 0,  # rf
               1,0,0, 0,0,0,1, 0, 0,  # r
               0,0,1, 0,0,0,1, 0, 0,  # p_E
               0,0,0, 0,0,0,0, 0, 0,  # epsilon
               0,0,0, 0,0,0,0, 0, 0,  # \theta^T
               0,0,0, 0,0,0,0, 0, 0),  # Omega^W
               9, 9, byrow=TRUE)

# Create adjacency matrix from transpose of auxiliary Jacobian and add column names
A_mat=t(M_mat)

# Create directed graph from adjacency matrix
library(igraph)
dg=graph_from_adjacency_matrix(A_mat, mode="directed", weighted= NULL)

# Define node labels
V(dg)$name=c(expression(Omega), "w", "p", expression(r^T), "r",
             expression(p_E), expression(epsilon), expression(theta^T), expression(Omega^W))


# Plot directed graph matrix
plot(dg, main="Directed graph of conflict inflation model", vertex.size=25, vertex.color="lightblue", 
     vertex.label.color="black", edge.arrow.size=0.4, edge.width=1.1, edge.size=2,
     edge.arrow.width=1.2, edge.color="black", vertex.label.cex=1, 
     vertex.frame.color="NA", margin=-0.08)
```

::: {.callout-note collapse="true" title="Python code"}
```{python}
#| echo: true
#| output: false
#| collapse: true

## Directed graph
import networkx as nx

# Construct auxiliary Jacobian matrix for 9 variables
# endogenous: (1) Omega, (2) w, (3) p, (4) r^T, (5) r, (6) p_E
# exogenous: (7) epsilon, (8) \theta^T, (9) Omega^W
M_mat = np.array([
    [0, 1, 1, 0, 0, 0, 0, 0, 0],  # Omega
    [1, 0, 0, 0, 0, 0, 0, 0, 1],  # w
    [0, 0, 0, 1, 1, 0, 0, 0, 0],  # p
    [0, 0, 0, 0, 0, 0, 0, 1, 0],  # r^T
    [1, 0, 0, 0, 0, 0, 1, 0, 0],  # r
    [0, 0, 1, 0, 0, 0, 1, 0, 0],  # p_E
    [0, 0, 0, 0, 0, 0, 0, 0, 0],  # epsilon
    [0, 0, 0, 0, 0, 0, 0, 0, 0],  # \theta^T
    [0, 0, 0, 0, 0, 0, 0, 0, 0]   # Omega^W
])

# Create adjacency matrix by transposing the auxiliary Jacobian matrix
A_mat = M_mat.T

# Create a directed graph from the adjacency matrix using networkx
G = nx.DiGraph(A_mat)

# Define node labels
nodelabs = {0: r'$\Omega$', 1: "w", 2: "p", 3: r'$r^T$', 4: "r", 5: r'$p^E$',
            6: r'$\epsilon$', 7: r'$\theta^T$', 8: r'$\Omega^W$'}

# Plot the graph
pos = nx.spring_layout(G, k=0.6)
nx.draw_networkx(G, pos, node_size=200, node_color="lightblue", 
                 edge_color="black", width=1.2, arrowsize=10, 
                 arrowstyle='->', font_size=8, font_color="black",
                 with_labels=True, labels=nodelabs)

plt.title("Directed graph of conflict inflation model", fontsize=12)
plt.show()

```
:::

The key exogenous variables in this version of the conflict inflation
model are the real energy price $\epsilon$, the target markup $\theta^T$, and
the target real wage of workers $\Omega^W$. The first two exogenous
variables feed into the endogenous conflict inflation process via their
impact on the realised profit margin $r$ and the desired profit margin
$r^T$, respectively. The gap between desired and realised profit
margins, in turn, impact price setting, the real wage, and then wage
setting. By contrast, the target real wage of workers enters the
inflationary process via its direct impact on nominal wages, which then
impacts real wages, the realised profit margin, and then price setting.
The directed graph thus highlights the joint determination and
endogenous interaction of prices, wages, and income distribution.

## Analytical discussion

To study the properties of the model, we reduce it to a first-order
dynamic system in $w_t$ and $p_t$. To this end, substitute
@eq-conflict_omega into @eq-conflict_w to obtain the equation for $w_t$.
Next, substitute @eq-conflict_r, @eq-conflict_rf and @eq-conflict_omega
into @eq-conflict_p. This gives the system:

$$
w_{t+1} = w_{t}\left[1+ \phi\left(\Omega^W - \frac{w_t}{p_t}\right)\right] \\
$$ {#eq-conflict_w_reduc} $$
p_{t+1} = p_{t} \left[1+\psi\left(\frac{w_t}{p_t} - \frac{1}{1+\theta^T} + \delta \epsilon \right) \right].
$$ {#eq-conflict_p_reduc}

Next, let's introduce the definitions
$\frac{w_{t+1} -w_{t}}{w_{t}} = \hat{w_t}$ and
$\frac{p_{t+1} -p_{t}}{p_{t}} = \hat{p_t}$ for the growth rates of wages
and prices, respectively. The equilibrium in this model is given by a
real wage and inflation rate $(\Omega^*, \hat{p}^*)$ for which
$\hat{w_t}=\hat{p_t}$, i.e. wages and prices grow at the same rate, such
that the real wage is constant. Imposing this equilibrium condition
using @eq-conflict_w_reduc and @eq-conflict_p_reduc yields:

$$\Omega^* = \frac{\phi \Omega^W+ \psi \left(\frac{1}{1+\theta^T}- \delta \epsilon \right)}{\phi + \psi}
$$ {#eq-conflict_omega_eq}

and

$$
\hat{p}^* = \phi \psi \left[\frac{ \Omega^W - \left(\frac{1}{1+\theta^T}- \delta \epsilon \right)}{\phi + \psi}\right],
$$ {#eq-conflict_infl_eq}

where the term $\left(\frac{1}{1+\theta^T}- \delta \epsilon\right)$ can also
be interpreted as firms' implicit real wage target $\Omega^F$, such that
for a positive equilibrium inflation rate $\hat{p}^* >0$ , we need
$\Omega^W > \Omega^F = \frac{1}{1+\theta^T}- \delta \epsilon$. Thus, workers'
and firms' real wage targets need to be mutually inconsistent for
inflation to emerge in equilibrium.

With these equilibrium solutions, we can show that increases
in $\theta^T$ and $\epsilon$ reduce the equilibrium real wage and raise the
equilibrium inflation rate, whereas increases in $\Omega^W$ raise the
equilibrium real wage and raise the inflation rate:

$$
\frac{\partial \Omega^*} {\partial \theta^T} = -\frac{\psi}{(\psi + \phi)(1+\theta^T)^2} < 0,
$$
$$
\frac{\partial \hat{p}^*}{\partial \theta^T} = \frac{\phi\psi}{(\psi + \phi)(1+\theta^T)^2} > 0,
$$

$$
\frac{\partial \Omega^*}{\partial \epsilon} = -\frac{\psi \delta}{\psi + \phi} < 0,
$$
$$
\frac{\partial \hat{p}^*}{\partial \epsilon} = \frac{\psi \phi \delta}{\psi + \phi} > 0,
$$


$$
\frac{\partial \Omega^*} {\partial \Omega^W} = \frac{\phi}{\psi + \phi} > 0,
$$
$$
\frac{\partial \hat{p}^*}{\partial \Omega^W} = \frac{\phi\psi}{\psi + \phi} > 0.
$$

Using the result for $\frac{\partial \Omega^*}{\partial \epsilon}$, we can easily work out the effects on the different measure of profitability listed in Table 2:

$$
\frac{\partial r^E}{\partial \epsilon} = \frac{\Omega^E}{\epsilon^2} > 0,
$$

$$
\frac{\partial r}{\partial \epsilon} =  -\frac{\delta \phi}{\psi + \phi} < 0,
$$

$$
\frac{\partial \pi}{\partial \epsilon} =   \frac{\delta(\psi r^T-\phi\Omega^W)}{(\psi + \phi)(1-\epsilon\delta)^2} \lesseqgtr 0, 
$$

$$
\frac{\partial \pi^{Total}}{\partial \epsilon} =  \frac{\delta \psi }{\psi + \phi} > 0.
$$

Next, to analyse dynamics of the model, we write down the Jacobian matrix of the system in
@eq-conflict_w_reduc and @eq-conflict_p_reduc:

$$
J(w,p)=\begin{bmatrix}
            1 + \hat{w} - \phi \Omega & \phi {\Omega}^2   \\
            \psi & 1 + \hat{p} - \psi \Omega   \\
        \end{bmatrix},
$$ where we used $\hat{w}=\phi\left(\Omega^W - \frac{w_t}{p_t}\right)$
and
$\hat{p}=\psi\left(\frac{w_t}{p_t} - \frac{1}{1+\theta^T} + \delta \epsilon \right)$
to simplify the expressions inside the Jacobian.

The stability conditions for two-dimensional systems in discrete time
are:

$$
1+tr(J)+det(J)>0,
$$ $$
1-tr(J)+det(J)>0,
$$ 
$$
1-det(J)>0,
$$ where $tr(J)$ is the trace and $det(J)$ is the determinant of the
Jacobian.

We know that this system will be unstable, because prices and wages will
grow exponentially in equilibrium. We thus want to derive the condition
under which the system becomes unstable and generates conflict
inflation.

Imposing $\hat{w_t}=\hat{p_t}$, the trace and determinant of the
Jacobian matrix in equilibrium are given by:

$$
tr(J)=2(1+\hat{w}) - \Omega(\phi+\psi),
$$


```{=tex}
\begin{equation}

\det(J)=1-\Omega(\phi+\psi)+\hat{w}[2+\hat{w} - \Omega(\phi+\psi)].

\end{equation}

```



Let's focus on the second stability condition, which is:

$$
1-tr(J)+det(J)= \hat{w}[\hat{w} - \Omega(\phi+\psi)]>0.
$$

Substituting the equilibrium solutions for $\hat{w}^*$ and $\omega^*$
into the term insight the square brackets yields:

$$
-\hat{w}^*\left(\frac{\omega^W\phi^2+(2\phi+\psi)\psi \Omega^F}{\phi+\psi}\right)<0.
$$

The term inside the brackets will be positive for meaningful economic
values. The term outside the brackets will be negative if
$\hat{w}^*=\hat{p^*}>0$, i.e. there is positive inflation, which in turn
requires $\Omega^W > \Omega^F$, i.e. workers' real wage target needs to
exceed firms' real wage target.

We can confirm these analytical results numerically. First, let's
confirm the solutions for the equilibrium real and the equilibrium
inflation rate as well as the 'instability condition' that the target
real wage of workers needs to exceed the target real wage of firms.

```{r conflict_9}
## Check solution for equilibrium real wage
# Analytical
(omega_eq= (phi*omega_w[1,1]+psi*(1/(1+theta_T[1,1]) -delta*epsilon[1,1]))/(phi+psi))
# Numerical
omega[1,Q]

## Check solution for equilibrium inflation rate
# Analytical
phi*psi*(omega_w[1,1]-(1/(1+theta_T[1,1]) -delta*epsilon[1,1]))/(phi+psi)
# Numerical
p_hat[1,Q]

## Check condition for instability: Omega^W > Omega^F
print(omega_w[1,1] > 1/(1+theta_T[1,1]) -delta*epsilon[1,1])
```

Next, we compute the Jacobian matrix of the simulated model (for the
baseline), obtain the eigenvalues, eigenvectors, trace, and determinant.
We then check the stability conditions. We find that the second
stability condition is indeed violated.

```{r conflict_10}
## Check eigenvalues of Jacobian matrix
# Define matrix
J=matrix(c(1+p_hat[1,Q]-phi*omega[1,Q], phi*omega[1,Q]^2,
           psi, 1+p_hat[1,Q]-psi*omega[1,Q]), 
           2, 2, byrow=TRUE)

# Obtain eigenvalues 
ev=eigen(J)
(evals = ev$values)
evecs = ev$vectors

# Obtain determinant and trace
det=det(J)      # determinant
tr=sum(diag(J)) # trace

#Check stability conditions
print(1+tr+det>0)
print(1-tr+det>0)
print(1-det>0)

```

Finally, following the approach presented in @sec-stab_analysis, we can
check whether the ratio of the two unstable state variables $w_t$ and
$p_t$, which is the real wage $\Omega$, approaches a value that is given
by the ratios of the elements in the dominant eigenvector. To do this,
we first normalise the eigenvectors by dividing through by the second
element (which corresponds to the second state variable in our system,
$p_t$). The first element in the normalised eigenvector should then
correspond to the equilibrium real wage. This is indeed the case.

```{r conflict_11}
# Normalise eigenvectors by dividing through by the second element
evecs_norm=evecs
for (i in 1:2){
  evecs_norm[,i]=evecs[,i]/evecs[2,i]
}

# Compare first element of normalised eigenvector with real wage Omega=w/p
evecs_norm[1,1]
omega[1,Q] 

```

::: {.callout-note collapse="true" title="Python code"}
```{python}
#| echo: true
#| output: false
#| collapse: true


### Analytical solutions
# Analytical solution for equilibrium real wage
print((phi * omega_w[0, 0] + psi * (1 / (1 + theta_T[0, 0]) - delta * epsilon[0, 0])) / (phi + psi))

# Numerical solution for equilibrium real wage
print(omega[0, Q-1])

# Analytical solution for equilibrium inflation rate
print(phi * psi * (omega_w[0, 0] - (1 / (1 + theta_T[0, 0]) - delta * epsilon[0, 0])) / (phi + psi))

# Numerical solution for equilibrium inflation rate
print(p_hat[0, Q-1]) 

# Check condition for instability: Omega^W > Omega*
print( omega_w[0, 0] > (1 / (1 + theta_T[0, 0]) - delta * epsilon[0, 0]))

# Define the Jacobian matrix J
J = np.array([
    [1 + p_hat[0, Q-1] - phi * omega[0, Q-1], phi * omega[0, Q-1]**2],
    [psi, 1 + p_hat[0, Q-1] - psi * omega[0, Q-1]]
])

# Obtain eigenvalues and eigenvectors
ev = np.linalg.eig(J)
evals = ev[0]  # eigenvalues
evecs = ev[1]  # eigenvectors
print(evals)
print(evecs)

# Obtain determinant and trace
det = np.linalg.det(J)
tr = np.trace(J)
print(det)
print(tr)

# Check stability conditions
print(1 + tr + det > 0)
print(1 - tr + det > 0)
print(1 - det > 0)

# Normalise eigenvectors by dividing through by the second element
evecs_norm = evecs.copy()  # Make a copy to modify
for i in range(2):
    evecs_norm[:, i] = evecs[:, i] / evecs[1, i]  # Normalize by the second element

# Compare first element of normalized eigenvector with real wage Omega = w/p
print(evecs_norm[0, 0])
print(omega[0, Q-1])

```
:::


## References
