# A Post-Kaleckian Distribution and Growth Model

## Overview

The post-Kaleckian growth model was developed by @bhaduri_unemployment_1990 and others to synthesise Marxian and Keynesian ideas about the effects of income distribution on economic growth.^[See @hein_distribution_2014, chap. 7 and @lavoie_post-keynesian_2014, chap. 6 for detailed treatments.] According to the Marxian view, capital accumulation is driven by profits. By contrast, Michal Kalecki and post-Keynesians such as Nicholas Kaldor argued that a redistribution of income towards profit-earners is likely to reduce consumption, as workers tend to have a higher marginal propensity to consume than capital owners. The post-Kaleckian growth model integrates these two mechanisms in a Keynesian framework in which aggregate demand and growth are demand-determined. It allows for wage-led as well as profit-led demand and growth regimes. In a wage-led regime, a redistribution of income towards workers has expansionary effects on aggregate demand (and possibly growth) as the expansionary effect on consumption outweighs the negative effect on investment. In a profit-led regime, the effect is contradictory as investment falls by more than consumption rises. Whether a regime is wage- or profit-led depends on the relative size of the propensities to consume and the propensity to invest.     

This is a model of long-run steady state growth. In the steady state, all endogenous variables grow at the same rate.^[All variables are normalised by the capital stock and thus rendered stationary.] Changes in parameters or exogenous variables lead to an instantaneous adjustment of the model's variables, so that the model can be analysed like a static one. We consider a version of the model with linear functions based on @hein_distribution_2014, chap. 7.2.2.

## The Model


$$
r=h \frac{u}{v}
$$ {#eq-r}

$$
s=[s_W + (s_\Pi - s_W)h]\frac{u}{v}, \quad 0 \geq s_W > s_\Pi \geq 1
$$ {#eq-s}

$$
c= \frac{u}{v}-s
$$ {#eq-c}

$$
g=g_0+g_1u+g_2h, \quad g_i > 0
$$ {#eq-g}

$$
u=v(c+g)
$$ {#eq-u}

where $r$, $s$, $c$, $g$, and $u$ are the profit rate, the saving rate, the consumption rate, the investment rate, and the rate of capacity utilisation, respectively.

@eq-r decomposes the profit rate into the product of the profit share $h$ (total profits over total output), the rate of capacity utilisation (actual output over potential output), and the inverse of $v$ (the capital-potential output ratio). Let $Y$ be output, $K$ be the capital stock, and $Y^P$ be potential output, the decomposition can also be written as $r=\frac{\Pi}{K}=\frac{\Pi}{Y}\frac{Y}{Y^P}\frac{Y^P}{K}$. The profit share and the capital-potential output ratio are taken to be exogenous in this model. Note also that the wage share is given by $1-h$. By @eq-s, the economy-wide saving rate is given by the sum of saving out of wages ($s_W(1-h)\frac{u}{v}$) and saving out of profits ($s_\Pi h\frac{u}{v}$). It is assumed that workers have a higher marginal propensity to consume than capital owners ($s_W>s_\Pi$). @eq-c simply states that consumption is income not saved. According to @eq-c, investment is determined by an autonomous component $g_0$ that may capture Keynesian `animal spirits', by the rate of capacity utilisation, and by the profit share. While the rate of capacity utilisation is a signal of (future) demand, the profit share may stimulate investment as internal funds are typically the cheapest source of finance. Finally, @eq-u is an equilibrium condition that assumes that the rate of capacity utilisation adjusts to clear the goods market.

The key question addressed by this model is a how a change in the profit share affects the rate of capacity utilisation and the rate of growth.^[@bhaduri_unemployment_1990 further discuss the effects on the profit rate.] As shown formally in the Appendix, there is no unambiguous answer to this question as the model encompasses different regimes. Three main regimes can be identified. First, a regime in which both the rate of utilisation and the rate of growth are negatively affected by an increase in the profit share. We will call this a wage-led demand and growth regime (WLD/WLG). Second, a regime in which the rate of utilisation is negatively affected and the rate of growth is positively affected by an increase in the profit share. We will call this a wage-led demand regime and profit-led growth regime (WLD/PLG). Third, a regime in which both the rate of utilisation and the rate of growth are positively affected by an increase in the profit share. We will call this a profit-led demand and growth regime (PLD/PLG).

## Simulation

### Parameterisation

Table 1 reports the parameterisation used in the simulation. We will consider three different parameterisations that represent the three regimes outlined above. For each of these regimes, there is a baseline scenario and a scenario in which the profit share ($h$) rises. This allows to assess the effects on the model's endogenous variables for the different regimes. 

**Table 1: Parameterisation**

| Scenario                        | $v$ | $s_W$ | $s_\Pi$ | $g_0$ | $g_1$ | $g_2$ | $h$ |
|:-------------------------------:|:---:|:-----:|:-------:|:-----:|:-----:|:-----:|:---:|
| 1a: baseline WLD/WLG            |  3  |  0.3  |   0.9   |  0.02 |  0.1  |  0.1  | 0.2 |
| 1b: rise in profit share ($h$)  |  3  |  0.3  |   0.9   |  0.02 |  0.1  |  0.1  | 0.3 |
| 2a: baseline WLD/PLG            |  3  |  0.3  |   0.9   |  0.02 |  0.08 |  0.1  | 0.2 |
| 2b: rise in profit share ($h$)  |  3  |  0.3  |   0.9   |  0.02 |  0.08 |  0.1  | 0.3 |
| 3a: baseline PLD/PLG            |  3  |  0.3  |   0.9   | -0.01 |  0.1  |  0.1  | 0.2 |
| 3b: rise in profit share ($h$)  |  3  |  0.3  |   0.9   | -0.01 |  0.1  |  0.1  | 0.3 |

### Simulation code

```{r pkdg}

#Clear the environment
rm(list=ls(all=TRUE))

# Set number of scenarios (including baselines)
S=6

#Create vector in which equilibrium solutions from different parameterisations will be stored
u_star=vector(length=S) # utilisation rate
g_star=vector(length=S) # growth rate of capital stock
s_star=vector(length=S) # saving rate
c_star=vector(length=S) # consumption rate
r_star=vector(length=S) # profit rate

# Set exogenous variables whose parameterisation changes across regimes 
g0=vector(length=S) # animal spirits
sw=vector(length=S) # propensity to save out of wages 
h=vector(length=S)  # profit share
g1=vector(length=S) # sensitivity of investment with respect to utilisation

### Construct different scenarios across 3 regimes: (1) WLD/WLG, (2) WLD/PLG, (3) PLD/PLG 

# baseline WLD/WLG
g0[1]=0.02
g1[1]=0.1
h[1]=0.2

# increase in profit share in WLD/WLG regime
g0[2]=0.02
g1[2]=0.1
h[2]=0.3

# baseline WLD/PLG
g0[3]=0.02
g1[3]=0.08
h[3]=0.2

# increase in profit share in WLD/PLG regime
g0[4]=0.02
g1[4]=0.08
h[4]=0.3
  
# baseline PLD/PLG
g0[5]=-0.01
g1[5]=0.1
h[5]=0.2

# increase in profit share in PLD/PLG regime
g0[6]=-0.01
g1[6]=0.1
h[6]=0.3

#Set constant parameter values
v=3    # capital-to-potential output ratio
g2=0.1 # sensitivity of investment with respect to profit share
sp=0.9 # propensity to save out of profits
sw=0.3 # propensity to save out of wages

#Check Keynesian stability condition for all scenarios
for (i in 1:S){
print(((sw+(sp-sw)*h[i])*(1/v) -g1[i])>0)
}

# Check demand and growth regime for 3 baseline scenarios
for (i in c(1,3,5)){
print(paste("Parameterisation", i, "yields:"))   
if(g2*(sw/v - g1[i])-g0[i]*(sp-sw)/v<0){
  print("wage-led demand regime")
  } else{
   print("profit-led demand regime")
  }
if(g1[i]*(g2*(sw/v - g1[i])-g0[i]*(sp-sw)/v)+g2*(((sw+(sp-sw)*h[i])*v^(-1)-g1[i])^2)<0){
  print("wage-led growth regime")
  } else{
  print("profit-led growth regime")
  }  
} 

# Initialise endogenous variables at some arbitrary positive value 
g=1
r=1
c=1
u=1
s=1

#Solve this system numerically through 1000 iterations based on the initialisation
for (i in 1:S){
  
  for (iterations in 1:1000){
    
    #(1) Profit rate
    r = (h[i]*u)/v
    
    #(2) Saving
    s = (sw+(sp-sw)*h[i])*(u/v)
    
    #(3) Consumption
    c= u/v-s
    
    #(4) Investment
    g = g0[i]+g1[i]*u+g2*h[i]
    
    #(5) Rate of capacity utilisation
    u = v*(c+g)
  }
  
  #Save results for different parameterisations in vector
  u_star[i]=u
  g_star[i]=g
  r_star[i]=r
  s_star[i]=s
  c_star[i]=c
}

```

### Plots

Figures @fig-utilisation-pkdg - @fig-profit-rate-pkdg depict the response of the model's key endogenous variables to changes in the profit share. In the first case of a wage-led demand and growth regime (WLD/WLG), investment is equally sensitive to a change in the rate of capacity utilisation ($g_1$) and a change in the profit share ($g_2$). A rise in the profit share reduces consumption, which reduces the rate of capacity utilisation and the rate of growth. This is despite a positive effect on the profit rate.^[If the negative effect on the rate of capacity utilisation was stronger, the profit rate could fall as well. See the Appendix for a formal derivation of the condition under which this may happen.] 

```{r pkdg2}
#| label: fig-utilisation-pkdg
#| fig-cap: Rate of capacity utilisation

barplot(u_star, ylab="u", names.arg=c("1a:baseline WLD/WLG", "1b:rise prof share", "2a:baseline WLD/PLG", "2b:rise prof share", "3a:baseline PLD/PLG", "3b: rise prof share"))

```

In the second case where the demand regime is wage-led but the growth regime is profit-led (WLD/PLG), investment is slightly less sensitive to a change in the rate of capacity utilisation compared to a change in the profit share. The rise in the profit share reduces consumption and the rate of utilisation, but the ultimate effect on investment is positive because investment reacts more strongly to the rise in the profit share than to the fall in demand.

```{r pkdg3}
#| label: fig-growth-pkdg
#| fig-cap: Rate of growth

barplot(g_star, ylab="g", names.arg=c("1:baseline WLD/WLG", "2:rise prof share", "3:baseline WLD/PLG",
                                      "4:rise prof share", 
                                      "5:baseline PLD/PLG", "6: rise prof share"))

```

Finally, in the third case where the demand regime and the growth regime are profit-led, investment is again equally sensitive to a change in the rate of capacity utilisation and to a change in the profit share, but now animal spirits are negative. A rise in the profit share now has strong positive effects on investment, which raises the rate of capacity utilisation and consumption.

```{r pkdg4}
#| label: fig-consumption-pkdg
#| fig-cap: Rate of consumption

barplot(c_star, ylab="c", names.arg=c("1:baseline WLD/WLG", "2:rise prof share", "3:baseline WLD/PLG",
                                      "4:rise prof share", 
                                      "5:baseline PLD/PLG", "6: rise prof share"))

```


```{r pkdg5}
#| label: fig-profit-rate-pkdg
#| fig-cap: Rate of profit

barplot(r_star, ylab="r", names.arg=c("1:baseline WLD/WLG", "2:rise prof share", "3:baseline WLD/PLG",
                                     "4:rise prof share", 
                                     "5:baseline PLD/PLG", "6: rise prof share"))

```

## Directed graph

Another perspective on the model's properties is provided by its directed graph. A directed graph consists of a set of nodes that represent the variables of the model. Nodes are connected by directed edges. An edge directed from a node $x_1$ to node $x_2$ indicates a causal impact of $x_1$ on $x_2$. 

```{r pkdgDAG}
#| message: false
#| label: fig-dg-pkdg
#| fig-cap: Directed graph of post-Kaleckian growth model

## Create directed graph 
# Construct auxiliary Jacobian matrix for 6 variables: 
# r, h, u, s, c, g

M_mat=matrix(c(0,1,1,0,0,0,
               0,0,0,0,0,0,
               0,0,0,0,1,1,
               0,1,1,0,0,0,
               0,0,1,1,0,0,
               0,1,1,0,0,0), 6, 6, byrow=TRUE)

# Create adjacency matrix from transpose of auxiliary Jacobian and add column names
A_mat=t(M_mat)
colnames(A_mat) = c("r", "h", "u", "s", "c", "g")

# Create and plot directed graph from adjacency matrix
library(igraph)
dg= graph_from_adjacency_matrix(A_mat, mode="directed", weighted= NULL)
plot(dg, main="", vertex.size=15, vertex.color="lightblue", 
     vertex.label.color="black", edge.arrow.size=0.4, edge.width=1.3,
     edge.arrow.width=0.5, edge.color="black", vertex.label.cex=1.2, 
     vertex.frame.color="NA", margin=0.01)

```

In Figure @fig-dg-pkdg, it can be seen that the profit share ($h$) is the key exogenous variable of the model.^[Other important exogenous variables or parameters that may shift but are not depicted here are animal spirits ($g_0$) or the saving propensities ($s_W, s_\Pi$). See @hein_distribution_2014, chap. 7.2.2, for a detailed discussion of their effects.] Consumption ($c$), saving ($s$), investment ($g$), and the rate of utilisation ($u$) form a closed loop (or cycle) within the system. The profit share affects both saving and investment, which in turn affect consumption and the rate of capacity utilisation. The profit rate is a residual variable (also called a `sink') in this model.

## Calculate analytical solutions

```{r pkdgsolution}

# Utilisation rate
for (i in 1:S){
print((g0[i]+g2*h[i])/((sw+(sp-sw)*h[i])/v-g1[i]))
}

# Growth rate
for (i in 1:S){
  print(((g0[i]+g2*h[i])*(sw+(sp-sw)*h[i])/v)/((sw+(sp-sw)*h[i])/v-g1[i]))
}

# Profit rate
for (i in 1:S){
  print((g0[i]+g2*h[i])*(h[i]/v)/((sw+(sp-sw)*h[i])/v-g1[i]))
}

```


## Analytical discussion

To find the equilibrium solutions, substitute @eq-s - @eq-g into @eq-u and solve for $u$:
\begin{align}
u^* = \frac{g_0+g_2h}{[s_W + (s_\Pi - s_W)h]v^{-1}-g_1}.
\end{align}

The equilibrium solution for $u$ can then be substituted into @eq-g to find:
\begin{align}
g^* = \frac{(g_0+g_2h)[s_W + (s_\Pi - s_W)h]v^{-1}}{[s_W + (s_\Pi - s_W)h]v^{-1}-g_1}.
\end{align}

The Keynesian stability condition requires $[s_W + (s_\Pi - s_W)h]v^{-1}-g_1>0$, i.e. saving need to react more strongly to income than investment.

The  equilibrium solution for $r$ can be found by substituting $u^*$ into @eq-r:
\begin{align}
r^* = \frac{h(g_0+g_2h)}{[s_W + (s_\Pi - s_W)h]-vg_1}.
\end{align}

To assess whether the demand regime is wage- or profit-led, take the derivative of $u^*$ with respect to $h$:
\begin{align}
\frac{\partial u^*}{\partial h} = \frac{\frac{s_W}{v}(g_0+g_2 )-(g_0\frac{s_\Pi}{v} + g_1g_2)}{[[s_W + (s_\Pi - s_W)h]v^{-1}-g_1]^2}.
\end{align}

It can be seen that, e.g., a higher propensity to save out of wages or negative animal spirits make the regime more likely to be profit-led.

By the same token, the sign of the derivative of $g^*$ with respect to $h$ determines whether the growth regime is wage- or profit-led:
\begin{align}
\frac{\partial g^*}{\partial h} = g_1\frac{\partial u^*}{\partial h}+g_2.
\end{align} 

It can be seen that, e.g., a higher sensitivity of investment with respect to the profit share makes the regime more likely to be profit-led.  

Finally, the effect on the profit rate will depend on the sign of the derivative:
\begin{align}
\frac{\partial r^*}{\partial h} = \frac{u^*}{v} + \frac{h}{v}\frac{\partial u^*}{\partial h},
\end{align}

which is likely to be positive but can become negative if the demand regime is strongly wage-led.

## References {-}